---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
# Unzip the original dataset and load it into a data.frame variable.
library(utils)
unzip("activity.zip", overwrite = FALSE)
activity <- read.csv("activity.csv", colClasses=c("numeric", "Date", "numeric"))
summary(activity)
```

## What is mean total number of steps taken per day?

```{r}
library(stats)
# Aggregate the data using the sum function in order to create the histogram.
total_steps_per_day <- aggregate(steps ~ date, activity, FUN = sum)
class(total_steps_per_day)
head(total_steps_per_day, n = 5)
```

### Histogram of the total number of steps taken each day

```{r histogram, fig.width=12}
library(ggplot2)
# Make histogram using ggplot2.
ggplot(total_steps_per_day, aes(x = factor(date), y = steps, fill = factor(date))) +
  geom_histogram(stat = "identity") + theme(legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1)) + labs(x = "Date",
      y = "Total steps", title = "Number of steps per day")
```

### Mean and median of the total number of steps taken per day

```{r}
steps_mean <- mean(total_steps_per_day$steps, na.rm = TRUE)
steps_median <- median(total_steps_per_day$steps, na.rm = TRUE)
```

The **mean** of the total steps per day is `r format(round(steps_mean, 2), nsmall = 2)` and the **median** is `r format(round(steps_median, 2), nsmall = 2)`.

## What is the average daily activity pattern?

```{r}
# Aggregate data and find position of maximum value of steps.
average_steps_per_interval <- aggregate(steps ~ interval, activity, FUN = mean)
max_average_steps_slot <- which.max(average_steps_per_interval$steps)
```

### Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.

```{r timeseries_1, fig.width=12}
plot(steps ~ interval, average_steps_per_interval, type = "l", main = "Daily activity pattern",
  xlab = "Interval", ylab = "Average no. of steps")
with(average_steps_per_interval, lines(x = interval[max_average_steps_slot], y = steps[max_average_steps_slot],
  type = "o", pch = 20, col = "red"))
with(average_steps_per_interval, text(x = interval[max_average_steps_slot], y= steps[max_average_steps_slot],
  labels = "Max", cex = 1, pos = 4, col = "red"))
```

### 5-minute interval containing the maximum number of steps.

```{r}
# Output the interval with the maximum average.
average_steps_per_interval[max_average_steps_slot, ]
```

## Imputing missing values

### Number of missing values per variable.

```{r}
# Output a summary of NAs per column.
sapply(activity, function(x) sum(is.na(x)))
```

### Imputing data using a strategy.

I will use the **average steps per interval** to impute the missing data, but will add some randomness too by calling the __runif__ function on each average. The use of a seed will allow making this strategy reproducible.

```{r}
# Function for calculating the number of steps based in the average.
imputableDataGenerator <- function(x) {
  set.seed(13)
  value <- round(runif(n = 1, min = x, max = x + 10))
  value
}

# This vector contains the value to imput for a given interval
# (e.g. interval == 0 --> steps == 9; interval == 20 --> steps == 7).
imputableData <- sapply(average_steps_per_interval[, 2], imputableDataGenerator)
average_steps_per_interval$imputableData <- imputableData

# Identify and keep the positions in the activity data.frame to fill.
positionsToFill <- which(is.na(x = activity$steps))

# Leave original DF untouched.
activity_copy <- activity

# Function that does the data imputation.
fillPositions <- function(positionsTofill) {
  for (i in positionsTofill) {
    theInt <- activity_copy[i, ]$interval
    whichInt <- which(average_steps_per_interval$interval == theInt)
    valueToInput <- average_steps_per_interval[whichInt, ]$imputableData
    activity_copy[i, ]$steps <<- valueToInput
  }
}

# Call function to fill missing data.
fillPositions(positionsToFill)
```

#### Comparing the resulting dataset to the previous one.

```{r}
# Original.
head(activity)

# With data imputed.
head(activity_copy)
```

### After data imputation

```{r}
total_steps_per_day_imputed <- aggregate(steps ~ date, activity_copy, FUN = sum)
```

### Histogram of the total number of steps taken each day

```{r histogram_after, fig.width=12}
# Make histogram using ggplot2.
ggplot(total_steps_per_day_imputed, aes(x = factor(date), y = steps, fill = factor(date))) +
  geom_histogram(stat = "identity") + theme(legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1)) + labs(x = "Date",
      y = "Total steps", title = "Number of steps per day after imputation")
```

### Mean and median of the total number of steps taken per day

```{r}
steps_mean_imputed <- mean(total_steps_per_day_imputed$steps, na.rm = TRUE)
steps_median_imputed <- median(total_steps_per_day_imputed$steps, na.rm = TRUE)
```

The **mean** of the total steps per day is `r format(round(steps_mean_imputed, 2), nsmall = 2)` and the **median** is `r format(round(steps_median_imputed, 2), nsmall = 2)`.

The impact of imputing data causes a slight difference in the average and median. Since my imputation strategy added some randomness and increased in ten units the number of steps (exactly 10, see **imputableDataGenerator**) to each missing observation, both values are higher.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Add day type column to imputed dataset.
activity_copy$daytype <- as.factor(ifelse(weekdays(activity_copy$date)
  %in% c("Saturday", "Sunday"), "Weekend", "Weekday")) 
```

### What is the average daily activity pattern, again?

```{r}
# Subset data for plotting.
weekend <- subset(activity_copy, subset = activity_copy$daytype == "Weekend")
average_steps_per_interval_weekend <- aggregate(steps ~ interval, weekend, FUN = mean)
weekday <- subset(activity_copy, subset = activity_copy$daytype == "Weekday")
average_steps_per_interval_weekday <- aggregate(steps ~ interval, weekday, FUN = mean)
```

### Time series plot of the 5-minute interval and the average number of steps taken, per day type.

```{r timeseries_2, fig.width=12, fig.height=12}
# Create a 2-row panel and add plots
par(mfrow = c(2, 1))
plot(steps ~ interval, average_steps_per_interval_weekend, type = "l", main = "Daily activity pattern weekends",
  xlab = "Interval", ylab = "Average no. of steps")
plot(steps ~ interval, average_steps_per_interval_weekday, type = "l", main = "Daily activity pattern weekdays",
  xlab = "Interval", ylab = "Average no. of steps")
```

In the plots there it can be noticed that people tends to walk **more** on weekends. The average is noticeably higher on saturdays and sundays.

Let's take summaries to confirm the previous affirmation:

```{r}
# Summarize data to help interpretation of plot.
summary(average_steps_per_interval_weekend)
summary(average_steps_per_interval_weekday)
```

- **Mean and median** for steps during weekends are: `r mean(average_steps_per_interval_weekend$steps)` and `r median(average_steps_per_interval_weekend$steps)`.
- **Mean and median** for steps during weekdays are: `r mean(average_steps_per_interval_weekday$steps)` and `r median(average_steps_per_interval_weekday$steps)`.
